GLM predictor Matrices

### 2020-12-23 scripts/analyze_trajectories_v2.Rmd
restructuting

```{r debugging}
# Debugging
args$burnin = 0.1
args$demes = read.csv("/Users/maceci/code/mt-analysis/201014_europe2/files/demes.csv", stringsAsFactors = FALSE)
args$input = "/Users/maceci/code/mt-analysis/201030_europe3/results/europe3c/201030_europe3.TL.traj"
args$metadata =  "/Users/maceci/code/mt-analysis/201014_europe2/results/europe2/final_metadata.tsv"
args$n = 50
demes = args$demes
demes$exclude_country[5] = "France,Germany,Italy,Spain"
demes$deme[3] = "China"
```

@@

# Data wrangling ---------------------------------------------------------------
states <- df$states %>%
  mutate(type = factor(type,
                       levels = 0:(nrow(demes) - 1),
                       labels = sort(demes$deme)),
         date_model = date(date_decimal(decimal_date(ymd(MRS)) - age))) %>%
  mutate(date = date_model + 10)

events <- df$events %>%
  mutate(src = factor(src,
                      levels = 0:(nrow(demes) - 1),
                      labels = sort(demes$deme)),
         dest = ifelse(is.na(dest), as.character(src),
                       as.character(factor(dest,
                                           levels = 0:(nrow(demes) - 1),
                                           labels = sort(demes$deme)))),
         date_model = date(date_decimal(decimal_date(ymd(MRS)) - age))) %>%
  mutate(date = date_model + 10)
```

@@

>>>>

```{r debugging}
# Debugging
args$burnin = 0.1
args$demes = read.csv("/Users/maceci/code/mt-analysis/201014_europe2/files/demes.csv", stringsAsFactors = FALSE)
args$input = "/Users/maceci/code/mt-analysis/europeGLM/results/europeC/201106_europe5.TL.traj"
args$metadata =  "/Users/maceci/code/mt-analysis/europeGLM/results/europeC/final_metadata.tsv"
args$n = 200
demes = args$demes
demes$exclude_country[5] = "France,Germany,Italy,Spain"
demes$deme[3] = "China"
demes$division[3] = NA
demes <- demes %>% arrange(deme)
```

@@

# Data wrangling ---------------------------------------------------------------
states <- df$states %>%
  mutate(type = factor(type,
                       levels = 0:(nrow(demes) - 1),
                       #labels = sort(demes$deme)),
                       labels = c("France", "Germany","China","Italy","OtherEuropean","Spain")),
         date_model = date(date_decimal(decimal_date(ymd(MRS)) - age))) %>%
  mutate(date = date_model + 10)

events <- df$events %>%
  mutate(src = factor(src,
                      levels = 0:(nrow(demes) - 1),
                      #labels = sort(demes$deme)),
                      labels = c("France", "Germany","China","Italy","OtherEuropean","Spain")),
         dest = ifelse(is.na(dest), as.character(src),
                       as.character(factor(dest,
                                           levels = 0:(nrow(demes) - 1),
                                           labels = sort(demes$deme)))),
         date_model = date(date_decimal(decimal_date(ymd(MRS)) - age))) %>%
  mutate(date = date_model + 10)
```

@@

<!-- ## First introduction -->
<!-- ```{r} -->
<!-- # 2. Events  -->
<!-- # 2.1. Time interval of first case (introduction) and first reported case to ECDC. -->

<!-- first_ecdc <- case_data %>%  -->
<!--   group_by(deme) %>%  -->
<!--   filter(cumcases != 0) %>%  -->
<!--   arrange(date) %>%  -->
<!--   slice(1)  %>%  -->
<!--   rename(dest = deme, min_date = date) -->

<!-- firstcase <- events %>% -->
<!--   filter(event == "M" | event == "O") %>% -->
<!--   mutate(dest = ifelse(is.na(dest), china_deme, dest)) %>% -->
<!--   group_by(dest, traj) %>% -->
<!--   arrange(time) %>% -->
<!--   slice(1) %>% -->
<!--   ungroup %>% -->
<!--   # Add row for Europe (all countries) -->
<!--   bind_rows(events %>% -->
<!--               filter(event == "M") %>% -->
<!--               group_by(traj) %>% -->
<!--               arrange(time) %>% -->
<!--               slice(1)  %>% -->
<!--               mutate(dest = "Europe")) -->

<!-- dates_firstcase <-  firstcase %>% -->
<!--   group_by(dest, min_date) %>% -->
<!--   summarize(Dmedian = median(date),  -->
<!--             Dlow    = quantile(date, 0.025, type = 1),  -->
<!--             Dhigh   = quantile(date, 0.975, type = 1), -->
<!--             .groups = "drop") %>% -->
<!--   mutate(difmedian = min_date - Dmedian, -->
<!--          diflow = min_date - Dlow, -->
<!--          difhigh = min_date - Dhigh)  -->

<!-- # 2.2. Time interval of first bith event (within region transmission) and first reported case -->
<!-- # to ECDC. -->
<!-- # How well did the countries detecting the first cases, there was already within region transmision? -->

<!-- # First introduction: Measure expected day of introduction. -->
<!-- firstbirth <- events %>% -->
<!--   filter(event == "B" | event == "O") %>% -->
<!--   group_by(dest, traj) %>% -->
<!--   arrange(time) %>% -->
<!--   slice(1) %>% -->
<!--   ungroup %>% -->
<!--   left_join(case_data %>% group_by(deme) %>% filter(cumcases != 0) %>% arrange(date) %>%slice(1)  %>% rename(dest = deme, min_date = date), by = "dest") %>% -->
<!--   bind_rows(events %>% -->
<!--               filter(dest != china_deme, event == "B") %>% -->
<!--               group_by(traj) %>% -->
<!--               arrange(time) %>% -->
<!--               slice(1)  %>% -->
<!--               mutate(dest = "Europe")) -->

<!-- dates_firstbirth <- firstbirth %>% -->
<!--   group_by(dest, min_date) %>% -->
<!--   summarize(Dmedian = median(date),  -->
<!--             Dlow    = quantile(date, 0.025, type = 1),  -->
<!--             Dhigh   = quantile(date, 0.975, type = 1), -->
<!--             .groups = "drop") %>% -->
<!--   mutate(difmedian = min_date - Dmedian, -->
<!--          diflow = min_date - Dlow, -->
<!--          difhigh = min_date - Dhigh) -->

<!-- p_firstbirth <- events %>% -->
<!--   filter(event == "B" | event == "O") %>% -->
<!--   bind_rows(events %>% -->
<!--               filter(dest != china_deme, event == "B") %>% -->
<!--               group_by(traj) %>% -->
<!--               arrange(time) %>% -->
<!--               slice(1)  %>% -->
<!--               mutate(dest = "Europe")) %>% -->
<!--   group_by(dest, traj) %>% -->
<!--   arrange(time) %>% -->
<!--   slice(1) %>% -->
<!--   ungroup %>% -->
<!--   left_join(case_data %>% group_by(deme) %>% filter(cumcases != 0) %>%  -->
<!--               arrange(date) %>% slice(1) %>% rename(dest = deme, min_date = date), by = "dest") %>% -->
<!--   replace_na(list(min_date = ymd("2020-01-25"))) %>% -->
<!--   filter(date <= min_date) %>% -->
<!--   count(dest) %>% -->
<!--   mutate(p = n/args$n * 100) -->

<!-- r_firstcases <- dates_firstcase%>% -->
<!--   left_join(dates_firstbirth, by = c("dest", "min_date")) %>% -->
<!--   left_join(p_firstbirth, by = "dest") %>% -->
<!--   replace_na(list(min_date = dates_firstcase$min_date[dates_firstcase$dest == "France"])) %>% -->
<!--   mutate(Country = dest, -->
<!--          `First Reported Case ECDC` = min_date, -->
<!--          `Median date (95% CCI) first introduction` = paste0(Dmedian.x, " (", Dlow.x, ", ", Dhigh.x, ")"), -->
<!--          `Median start date (95% CCI) within region transmission` = paste0(Dmedian.y, " (", Dlow.y, ", ", Dhigh.y, ")"), -->
<!--          `Percentage of trajectories within region transmission when first reported case` = paste0(p, "%")) %>% -->
<!--   select(c("Country", "First Reported Case ECDC", "Median date (95% CCI) first introduction", "Median start date (95% CCI) within region transmission", -->
<!--            "Percentage of trajectories within region transmission when first reported case")) -->

<!-- # 2.1. Time interval of first outcoming migration from a country compared to first reported case -->
<!-- # In which percentage of trajectories, transmission from that country would have been avoided if -->
<!-- # any ways of outcoming migrations were closed the day of first reported case. -->

<!-- firstmig <- events %>% -->
<!--   filter(event == "M") %>% -->
<!--   group_by(src, traj) %>% -->
<!--   arrange(time) %>% -->
<!--   slice(1) %>% -->
<!--   ungroup %>% -->
<!--   left_join(case_data %>% group_by(deme) %>% filter(cumcases != 0) %>% arrange(date) %>% slice(1)  %>% rename(src = deme, min_date = date), by = "src")  -->

<!-- dates_firstmig <- firstmig%>% -->
<!--   group_by(src, min_date) %>% -->
<!--   summarize(Dmedian = median(date),  -->
<!--             Dlow    = quantile(date, 0.025, type = 1),  -->
<!--             Dhigh   = quantile(date, 0.975, type = 1), -->
<!--             .groups = "drop") %>% -->
<!--   mutate(difmedian = min_date - Dmedian, -->
<!--          diflow = min_date - Dlow, -->
<!--          difhigh = min_date - Dhigh)  -->

<!-- p_firstmig <- events %>% -->
<!--   filter(event == "M") %>% -->
<!--   group_by(src, traj) %>% -->
<!--   arrange(time) %>% -->
<!--   slice(1) %>% -->
<!--   ungroup %>% -->
<!--   left_join(case_data %>% group_by(deme) %>% filter(cumcases != 0) %>% arrange(date) %>% slice(1)  %>% rename(src = deme, min_date = date), by = "src") %>% -->
<!--   filter(date >= min_date) %>% -->
<!--   count(src) %>% -->
<!--   mutate(p = n/args$n * 100) -->

<!-- r_firstmig <- dates_firstmig %>% -->
<!--   left_join(p_firstmig, by = "src") %>% -->
<!--   mutate(Country = src, -->
<!--          `First Reported Case ECDC` = min_date, -->
<!--          `Median date (95% CCI) first outgoing migration` = paste0(Dmedian, " (", Dlow, ", ", Dhigh, ")"), -->
<!--          `Percentage of trajectories first migration after first reported case` = paste0(p, "%")) %>% -->
<!--   select(-one_of(c(colnames(dates_firstmig), colnames(p_firstmig)))) -->

<!-- # Measure the time between the first migration into a country and the first birth in the country.  -->
<!-- # Or the first outcoming mirgration from the country and first birth.  -->
<!-- # This could be interesting to say if we should focus or not the screening and testing  -->
<!-- # capacities to detect incoming migrations or a more general strategy to find cases in  -->
<!-- # the population. Is it different for each country? -->

<!-- firsts <- firstcase %>% -->
<!--   select(dest, date, min_date) %>% -->
<!--   rename(deme = dest, introduction = date) %>% -->
<!--   left_join(firstbirth %>% select(date, dest) %>% rename(deme = dest, birth = date), by = "deme") %>% -->
<!--   left_join(firstmig %>% select(date, src) %>% rename(deme = src, mig = date), by = "deme") %>% -->
<!--   filter(deme != "Europe") %>% -->
<!--   mutate(dif1 = birth - introduction, -->
<!--          dif2 = mig - introduction) %>% -->
<!--   group_by(deme) %>% -->
<!--   summarize(dif1median = median(dif1),  -->
<!--             dif1low    = quantile(dif1, 0.025, type = 1),  -->
<!--             dif1high   = quantile(dif1, 0.975, type = 1), -->
<!--             dif2median = median(dif2),  -->
<!--             dif2low    = quantile(dif2, 0.025, type = 1),  -->
<!--             dif2high   = quantile(dif2, 0.975, type = 1), -->
<!--             .groups = "drop")  -->

<!-- r_firsts <- firsts %>% -->
<!--   mutate(Country = deme, -->
<!--          `Days between introduction and first within region transmision` = paste0(dif1median, " (", dif1low, ", ", dif1high, ")"), -->
<!--          `Days between introduction and first outgoing migration` = paste0(dif2median, " (", dif2low, ", ", dif2high, ")")) %>% -->
<!--   select(-one_of(colnames(firsts))) %>% -->
<!--   left_join(r_firstcases, by = "Country") %>% -->
<!--   left_join(r_firstmig, by = c("Country", "First Reported Case ECDC")) %>% -->
<!--   select(1,4,5,6,8,2,3,7,9) -->

<!-- write_tsv(r_firsts, paste0(args$output_table, "02.tsv")) -->

<!-- ``` -->

### 2020-12-23 scripts/predictor_matrices.Rmd
restructuting

---
title: "GLM Predictor Matrices"
subtitle: "Analyses Europe6 and Europe7"
author: "Cecilia Valenzuela"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output:
  beamer_presentation:
    theme: "Pittsburgh"
    colortheme: "dove"
    fonttheme: "structurebold"
    highlight: pygments
classoption: "aspectratio=169"
fontsize: 8pt
tables: yes
---
```{r, setup, include=FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

@@

$$
m_{ij}^{e} = c \exp(\beta_{pas} \delta_{pas} x_{ij.pas}^e + \beta_{pop} \delta_{pop }x_{ij.pop} + \beta_{dist} \delta_{dist}x_{ij.dist})
$$

where $\boldsymbol{\beta}$ are the coefficients for the predictors that can be between $-\infty$ and $\infty$ (prior  $\mathcal{N}(0,4)$) and $\boldsymbol{\delta}$ are the indicators that can be 0 or 1 and denote if a predictor contributes at all.

Analysis Europe 6 without indicator variables, Europe 7 with indicator variables, same predictors.

@@

Passengers carried data from Eurostats. Reported _departures_ values are taken for European countries (recommendation from Eurostats to avoid duplicates). For China, arrivals and departures values are considered.

@@

# These datasets are dowload from eurostats 
intraUE_data <- "./data/avia_paincc.tsv"
extraUE_data <- "./data/avia_paexcc.tsv"

flight_df <- get_flightData(intraUE_data, extraUE_data, countries = demes)

flight_df_3e <- flight_df %>%
  filter((MONTH %in% 1:3 & YEAR == 2020) | (MONTH == 12 & YEAR == 2019)) %>%
  mutate(EPOCH = case_when(
    MONTH == 2 ~ 2,
    MONTH == 3 ~ 3,
    TRUE ~ 1),
    PRED = 1) %>%
  rename(SRC = GEO, DEST = PARTNER) %>%
  group_by(SRC, DEST, EPOCH, PRED) %>%
  summarise(N = n(),
         EVALUE = sum(DVALUE/N)) %>%
  ungroup %>%
  # Transformations
  mutate(LVALUE = log(EVALUE + 1),
         SLVALUE = scale(LVALUE)) %>%
  arrange(EPOCH, factor(SRC, levels = demes), factor(DEST, levels = demes))

#write(slv_flightMatrix_3e, "files/flight_matrix_3e.csv", sep = ",")
```

```{r, echo = FALSE}
flight_matrices <- lapply(1:3, function (i) {
  flight_df_3e %>%
  filter(EPOCH == i) %>%
  select(SRC, DEST, SLVALUE) %>%
  pivot_wider(names_from = "DEST", values_from = "SLVALUE") %>%
  mutate_if(is.numeric, ~round(.,2)) %>%
  column_to_rownames("SRC") %>%
  select(all_of(unname(demes)))
})

@@

# Fligh data plot
ggplot(flight_df_3e, aes(x = EPOCH, y = EVALUE, color = DEST)) +
  geom_line() +
  geom_point(size = 1) +
  scale_y_log10(labels = trans_format("log10", math_format(10^.x)), breaks = trans_breaks('log10', function(x) 10^x)) +
  scale_x_continuous(breaks = c(1, 2, 3), labels = c("E1", "E2", "E3")) +
  facet_grid(~SRC) +
  xlab("Epoch") +
  ylab("# Daily Flight Passengers") + 
  #theme(axis.title.x = element_blank()) + 
  scale_color_manual(values = dcolors, name = "Destination")
```

@@

```{r, echo=FALSE}
data(pop)
population <- pop %>% filter(name %in% labels(demes)) %>% select(name, "2020") 
pop_eu_demes <- population %>% filter(name != "China") %>% summarise(s = sum(`2020`)) %>% pull(s)
pop_eu_total <- pop %>% filter(name == "Europe") %>% pull("2020")
population <- population %>%
  tibble::add_row(name = "OtherEuropean", `2020` = pop_eu_total - pop_eu_demes) %>%
  arrange(name) %>%
  rename(POP = 2) %>%
  mutate(SLVALUE = scale(log(POP)))
pop_df <- population %>%
  mutate(n = 6, SRC = demes) %>%
  uncount(n) %>%
  mutate(DEST = rep(demes, 6),
         POP = ifelse(SRC == DEST, NA, POP),
         SLVALUE = ifelse(SRC == DEST, NA, SLVALUE)) %>%
  select(-name) %>%
  filter(!is.na(POP))
pop_df_3e <- bind_rows(replicate(3, pop_df %>% mutate(PRED = 2), simplify = FALSE)) %>% 
  mutate(EPOCH = rep(1:3, each = 30, length = 90)) %>%
  bind_rows(bind_rows(replicate(3, pop_df %>% 
                        arrange(factor(DEST, levels = demes),
                                factor(SRC, levels = demes)) %>%
                          mutate(TEMP = SRC,
                                 SRC = DEST,
                                 DEST = TEMP,
                                 PRED = 3),
                      simplify = FALSE)) %>%
  mutate(EPOCH = rep(1:3, each = 30, length = 90)))
```

@@

\tiny
```{r, echo=FALSE}
pop_matrices <- lapply(c("DEST", "SRC"), function (x) {
  pop_df %>%
  select(-POP) %>%
  pivot_wider(names_from = all_of(x), values_from = "SLVALUE") %>%
  mutate_if(is.numeric, ~round(., 2)) %>%
  mutate_at(1, ~factor(., levels = demes)) %>%
  arrange_at(1) %>%
  select(1, all_of(unname(demes)))
})

@@

dist_df <- centroids %>%
  full_join(centroids, by = character())
dist_df <- dist_df%>%
  mutate(DIST = distHaversine(select(dist_df, longitude.x, latitude.x),
                              select(dist_df, longitude.y, latitude.y))) %>%
  group_by(deme.x, deme.y) %>%
  summarise(DIST = mean(DIST), .groups = "drop") %>%
  filter(deme.x != deme.y) %>%
  rename(SRC = deme.x, DEST = deme.y) %>% 
  mutate(SLVALUE = scale(log(DIST))) %>%
  arrange(factor(SRC, levels = demes), factor(DEST, levels = demes))
dist_df_3e <- bind_rows(replicate(3, dist_df, simplify = FALSE)) %>%
  mutate(EPOCH = rep(1:3, each = 30, length = 90), PRED = 4)
```

@@

\tiny
```{r, echo = FALSE}
dist_matrix <- dist_df %>%
  select(-DIST) %>%
  pivot_wider(names_from = "DEST", values_from = "SLVALUE") %>%
  mutate_if(is.numeric, ~round(., 2)) %>%
  column_to_rownames("SRC") %>%
  select(all_of(unname(demes)))

@@

@@

```{r, echo = FALSE}
pred_3e <- bind_rows(flight_df_3e, pop_df_3e) %>%
  bind_rows(dist_df_3e) %>%
  arrange(PRED, desc(EPOCH), factor(SRC, levels = demes), factor(DEST, levels = demes)) %>%
  select(PRED, EPOCH, SRC, DEST, SLVALUE)

@@

mpred_3e <- as.data.frame(matrix(pred_3e %>% pull(SLVALUE), ncol = 5, byrow = TRUE))
write_csv(mpred_3e, "./files/covariates_4p_3e.csv", col_names = FALSE)
```

@@

Covariate list saved as `files/covariates_4p_3e.csv`

>>>>

---
title: "GLM predictor Matrices"
subtitle: "Analyses Europe6 and Europe7"
author: "Cecilia Valenzuela"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output:
  beamer_presentation:
    theme: "Pittsburgh"
    colortheme: "dove"
    fonttheme: "structurebold"
    highlight: pygments
classoption: "aspectratio=169"
fontsize: 8pt
tables: yes
---
```{r, setup, include=FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

@@

$$
m_{ij}^{e} = c \exp(\beta_{pas} \delta_{pas} x_{ij.pas}^e + \beta_{popsrc} \delta_{popsrc}x_{ij.popsrc} + \beta_{popdes} \delta_{popdes}x_{ij.popdes} + \beta_{dist} \delta_{dist}x_{ij.dist})
$$

where $\boldsymbol{\beta}$ are the coefficients for the predictors that can be between $-\infty$ and $\infty$ (prior  $\mathcal{N}(0,4)$, random walk operator). The indicator $\boldsymbol{\delta}$ can be 0 or 1 and denote if a predictor contributes at all (Bit Flip Operator).

@@

Daily number of passengers betwwen demes. Passengers carried data from Eurostats. Reported _departures_ values are taken for European countries (recommendation from Eurostats to avoid duplicates). For China, arrivals and departures values are considered.

@@

flight_df <- get_flightData(countries = demes)

flight_df_3e <- flight_df %>%
  filter((month(time) %in% 1:3 & year(time) == 2020) | (month(time) == 12 & year(time) == 2019)) %>%
  mutate(epoch = case_when(
    month(time) == 2 ~ 2,
    month(time) == 3 ~ 3,
    TRUE ~ 1),
    pred = 1) %>%
  rename(src = geo, dest = partner) %>%
  group_by(src, dest, epoch, pred) %>%
  summarise(N = n(),
         evalues = sum(dvalues/N)) %>%
  ungroup %>%
  # Transformations
  mutate(lvalues = log(evalues + 1),
         slvalues = scale(lvalues)) %>%
  arrange(epoch, factor(src, levels = demes), factor(dest, levels = demes))
```

```{r, echo = FALSE}
flight_matrices <- lapply(1:3, function (i) {
  flight_df_3e %>%
  filter(epoch == i) %>%
  select(src, dest, slvalues) %>%
  pivot_wider(names_from = "dest", values_from = "slvalues") %>%
  mutate_if(is.numeric, ~round(.,2)) %>%
  column_to_rownames("src") %>%
  select(all_of(unname(demes)))
})

@@

# Fligh data plot
ggplot(flight_df_3e, aes(x = epoch, y = evalues, color = dest)) +
  geom_line() +
  geom_point(size = 1) +
  scale_y_log10(labels = trans_format("log10", math_format(10^.x)), breaks = trans_breaks('log10', function(x) 10^x)) +
  scale_x_continuous(breaks = c(1, 2, 3), labels = c("E1", "E2", "E3")) +
  facet_grid(~src) +
  xlab("Epoch") +
  ylab("# Daily Flight Passengers") + 
  #theme(axis.title.x = element_blank()) + 
  scale_color_manual(values = dcolors, name = "Destination")
```

@@

```{r, echo=FALSE}
data(pop)
population <- pop %>% filter(name %in% labels(demes)) %>% select(name, "2020") 
pop_eu_demes <- population %>% filter(name != "China") %>% summarise(s = sum(`2020`)) %>% pull(s)
pop_eu_total <- pop %>% filter(name == "Europe") %>% pull("2020")
population <- population %>%
  tibble::add_row(name = "OtherEuropean", `2020` = pop_eu_total - pop_eu_demes) %>%
  arrange(name) %>%
  rename(popvalue = 2) %>%
  mutate(slvalues = scale(log(popvalue)))
pop_df <- population %>%
  mutate(n = 6, src = demes) %>%
  uncount(n) %>%
  mutate(dest = rep(demes, 6),
         popvalue = ifelse(src == dest, NA, popvalue),
         slvalues = ifelse(src == dest, NA, slvalues)) %>%
  select(-name) %>%
  filter(!is.na(popvalue))
pop_df_3e <- bind_rows(replicate(3, pop_df %>% mutate(pred = 2), simplify = FALSE)) %>% 
  mutate(epoch = rep(1:3, each = 30, length = 90)) %>%
  bind_rows(bind_rows(replicate(3, pop_df %>% 
                        arrange(factor(dest, levels = demes),
                                factor(src, levels = demes)) %>%
                          mutate(tmp = src,
                                 src = dest,
                                 dest = tmp,
                                 pred = 3),
                      simplify = FALSE)) %>%
  mutate(epoch = rep(1:3, each = 30, length = 90)))
```

@@

\tiny
```{r, echo=FALSE}
pop_matrices <- lapply(c("dest", "src"), function (x) {
  pop_df %>%
  select(-popvalue) %>%
  pivot_wider(names_from = all_of(x), values_from = "slvalues") %>%
  mutate_if(is.numeric, ~round(., 2)) %>%
  mutate_at(1, ~factor(., levels = demes)) %>%
  arrange_at(1) %>%
  select(1, all_of(unname(demes)))
})

@@

dist_df <- centroids %>%
  full_join(centroids, by = character())
dist_df <- dist_df%>%
  mutate(dist = distHaversine(select(dist_df, longitude.x, latitude.x),
                              select(dist_df, longitude.y, latitude.y))) %>%
  group_by(deme.x, deme.y) %>%
  summarise(dist = mean(dist), .groups = "drop") %>%
  filter(deme.x != deme.y) %>%
  rename(src = deme.x, dest = deme.y) %>% 
  mutate(slvalues = scale(log(dist))) %>%
  arrange(factor(src, levels = demes), factor(dest, levels = demes))
dist_df_3e <- bind_rows(replicate(3, dist_df, simplify = FALSE)) %>%
  mutate(epoch = rep(1:3, each = 30, length = 90), pred = 4)
```

@@

\tiny
```{r, echo = FALSE}
dist_matrix <- dist_df %>%
  select(-dist) %>%
  pivot_wider(names_from = "dest", values_from = "slvalues") %>%
  mutate_if(is.numeric, ~round(., 2)) %>%
  column_to_rownames("src") %>%
  select(all_of(unname(demes)))

@@

## Daily flight passengers by source population

Daily number of passengers betwwen demes normalized by the population of the source deme. Passengers carried data from Eurostats. Reported _departures_ values are taken for European countries (recommendation from Eurostats to avoid duplicates). For China, arrivals and departures values are considered.

flight_df <- get_flightData(countries = demes)

flight_df_3e_n <- flight_df %>%
  filter((month(time) %in% 1:3 & year(time) == 2020) | (month(time) == 12 & year(time) == 2019)) %>%
  mutate(epoch = case_when(
    month(time) == 2 ~ 2,
    month(time) == 3 ~ 3,
    TRUE ~ 1),
    pred = 1) %>%
  rename(src = geo, dest = partner) %>%
  left_join(population %>% mutate(src = demes[name]), by = "src") %>%
  group_by(src, dest, epoch, pred, popvalue) %>%
  summarise(N = n(),
         evalues = sum(dvalues/N)) %>%
  mutate(nevalues = evalues/(popvalue*1000)) %>%
  ungroup %>%
  # Transformations
  mutate(#lvalues = log(nevalues + 1),
         slvalues = scale(nevalues)) %>%
  arrange(epoch, factor(src, levels = demes), factor(dest, levels = demes))
```

```{r, echo = FALSE}
flight_matrices <- lapply(1:3, function (i) {
  flight_df_3e_n %>%
  filter(epoch == i) %>%
  select(src, dest, slvalues) %>%
  pivot_wider(names_from = "dest", values_from = "slvalues") %>%
  mutate_if(is.numeric, ~round(.,2)) %>%
  column_to_rownames("src") %>%
  select(all_of(unname(demes)))
})

# Fligh data plot
ggplot(flight_df_3e_n, aes(x = epoch, y = nevalues, color = dest)) +
  geom_line() +
  geom_point(size = 1) +
  #scale_y_log10(labels = trans_format("log10", math_format(10^.x)), breaks = trans_breaks('log10', function(x) 10^x)) +
  scale_x_continuous(breaks = c(1, 2, 3), labels = c("E1", "E2", "E3")) +
  facet_grid(~src) +
  xlab("Epoch") +
  ylab("# Daily Flight Passengers") + 
  #theme(axis.title.x = element_blank()) + 
  scale_color_manual(values = dcolors, name = "Destination")
```

@@

```{r, echo = FALSE}
pred_3e <- bind_rows(flight_df_3e, pop_df_3e) %>%
  bind_rows(dist_df_3e) %>%
  arrange(pred, desc(epoch), factor(src, levels = demes), factor(dest, levels = demes)) %>%
  select(pred, epoch, src, dest, slvalues)

@@

mpred_3e <- as.data.frame(matrix(pred_3e %>% pull(slvalues), ncol = 5, byrow = TRUE))
write_csv(mpred_3e, "./files/covariates_4p_3e.csv", col_names = FALSE)
```

```{r, echo = FALSE}
pred_3e <- bind_rows(flight_df_3e_n, pop_df_3e) %>%
  bind_rows(dist_df_3e) %>%
  arrange(pred, desc(epoch), factor(src, levels = demes), factor(dest, levels = demes)) %>%
  select(pred, epoch, src, dest, slvalues)

mpred_3e <- as.data.frame(matrix(pred_3e %>% pull(slvalues), ncol = 5, byrow = TRUE))
write_csv(mpred_3e, "./files/covariates_4p_3e_n.csv", col_names = FALSE)
```

@@

Covariate list saved as `files/covariates_4p_3e.csv` and `files/covariates_4p_3e_n.csv` (normalized by population passenger data).

## Analyses

- **Europe 4:** GLM with daily flight data as predictor, constant over all the analysis time.
- **Europe 5:** GLM with daily flight data as predictor, 3 epochs.
- **Europe 5b:** GLM with daily flight data normalized by population of the deme, 3 epochs.
- **Europe 6** GLM with the 4 predictors described here, without indicator variables 
- **Europe 6b** GLM with the 4 predictors described here (fligh data normalized by population), without indicator variables 
- **Europe 7** with indicator variables, same predictors than eu6.
<!-- - **Europe 7b** with indicator variables, same predictors than eu7. -->
